---
title: "Function to raise number to cube in Fortran"
output:
  html_document: default
  html_notebook: default
---

## The Fortran function 
The factorial function resides in the file `cube.f`.
Here is the code.

```
      subroutine cube(n, x)

      integer n
      double precision x(n)
      integer i

      do 100 i = 1, n
        x(i) = x(i) ** 3
 100  continue

      end

```   

Notice that `cube` is a subroutine not a function, so, we will not need a Fortran wrapper like we did in `fib.f`.

## Compiling the file `cube` from the notebook
This could be done from the R prompt directly. In this case, we will link and compile to generate the DLL from the notebook.

Sometimes, the DLL file is locked because it was called by R but not released. If this happens, exit R and try deleting the files again.
```{r results="hold"}
if(is.loaded("cube")) dyn.unload(paste("cube", .Platform$dynlib.ext, sep = ""))
# delete existing files first
file.remove("cube.o")
file.remove("cube.dll")
file.remove("cube.so")

# now, compile
system("R CMD SHLIB cube.f")

```
The file `cube.dll` should be there now.


## Load the subroutine. Generic OS call
```{r}
# this loads the library in any system, Windows or Linux
dyn.load(paste("cube", .Platform$dynlib.ext, sep = ""))

# or use this
# dyn.load('cube.dll') for Windows only
```

We call this function with:
```{r}

cube <- function(x) {
  out <-  .Fortran("cube", 
                   n = as.integer(x), 
                   x = double(x))
  return(out)
}

# call the R function wrapper
cube(10)    
```


We can also call the Fortran function multiple times, like this:
```{r}
for(num in 1:20)
  cat(.Fortran("cube", 
                   n = as.integer(num), 
                   x = double(num))$x, "\n")
```

## Unload the Fortran subroutine. Generic OS
```{r}
dyn.unload(paste("cube", .Platform$dynlib.ext, sep = ""))
```


