---
title: "R Notebook"
output:
  html_document: default
  html_notebook: default
---

```{r setup, include=F, error=T, message=F, warning=F}
knitr::opts_chunk$set(echo=T, comment=NA, error=T, warning=F, message = F, fig.align = 'center')
```


## Description of the Fortran subroutine
The code of the main file an be found in `dgetrf.f`. It has other Fortran dependencies.

```
      SUBROUTINE DGETRF( M, N, A, LDA, IPIV, INFO )
*
*  -- LAPACK routine (version 3.2) --
*  -- LAPACK is a software package provided by Univ. of Tennessee,    --
*  -- Univ. of California Berkeley, Univ. of Colorado Denver and NAG Ltd..--
*     November 2006
*
*     .. Scalar Arguments ..
      INTEGER            INFO, LDA, M, N
*     ..
*     .. Array Arguments ..
      INTEGER            IPIV( * )
      DOUBLE PRECISION   A( LDA, * )
*     ..
*
*  Purpose
*  =======
*
*  DGETRF computes an LU factorization of a general M-by-N matrix A
*  using partial pivoting with row interchanges.
*
*  The factorization has the form
*     A = P * L * U
*  where P is a permutation matrix, L is lower triangular with unit
*  diagonal elements (lower trapezoidal if m > n), and U is upper
*  triangular (upper trapezoidal if m < n).
*
*  This is the right-looking Level 3 BLAS version of the algorithm.
*
*  Arguments
*  =========
*
*  M       (input) INTEGER
*          The number of rows of the matrix A.  M >= 0.
*
*  N       (input) INTEGER
*          The number of columns of the matrix A.  N >= 0.
*
*  A       (input/output) DOUBLE PRECISION array, dimension (LDA,N)
*          On entry, the M-by-N matrix to be factored.
*          On exit, the factors L and U from the factorization
*          A = P*L*U; the unit diagonal elements of L are not stored.
*
*  LDA     (input) INTEGER
*          The leading dimension of the array A.  LDA >= max(1,M).
*
*  IPIV    (output) INTEGER array, dimension (min(M,N))
*          The pivot indices; for 1 <= i <= min(M,N), row i of the
*          matrix was interchanged with row IPIV(i).
*
*  INFO    (output) INTEGER
*          = 0:  successful exit
*          < 0:  if INFO = -i, the i-th argument had an illegal value
*          > 0:  if INFO = i, U(i,i) is exactly zero. The factorization
*                has been completed, but the factor U is exactly
*                singular, and division by zero will occur if it is used
*                to solve a system of equations.
*
*  =====================================================================
```

## compiling several Fortran modules
Use `-fPIC` at the end.
```{r results='hold'}
if(is.loaded("dgetrf")) dyn.unload(paste("dgetrf", .Platform$dynlib.ext, sep = ""))
# delete existing files first
file.remove("dgetrf.o")
file.remove("dgetrf.dll")
file.remove("dgetrf.so")

# now, compile
system("R CMD SHLIB  dgetrf.f iparam2stage.F iparmq.f dgetf2.f dlaswp.f  dlamch.f  ilaenv.f lsame.f ieeeck.f xerbla.f dgemm.f dtrsm.f -fPIC")

```

## Load the library
```{r}
# this loads the library in any system, Windows or Linux
dyn.load(paste("dgetrf", .Platform$dynlib.ext, sep = ""))
```


## Call the Fortran library from R
To call `dgetrf` from R we use .Fortran.
```{r}
# Run DGETRF
mm = matrix(rnorm(16), nr = 4)       # input data

mmdc = .Fortran("dgetrf", 
                m = as.integer(4), n = as.integer(4),   # input
                a = as.double(mm),                      # input/output
                lda = as.integer(4),                    # input
                ipiv = integer(4), info = integer(1))   # output
str(mmdc)
```

The output should be similar to this:

```
List of 6
$ m : int 4
$ n : int 4
$ a : num [1:16] 1.391 0.141 -0.281 -0.106 -1.375 ...
$ lda : int 4
$ ipiv: int [1:4] 4 3 4 4
$ info: int 0
```

## The R function
* Be careful to match the argument types by explicit coersions with
`as.integer` or `as.double` or by creating vectors of an explicit
storage mode and length using the constructors `integer(n)` and
`double(n)`

```{r}
dgetrf <- function(m, n, mm = matrix(rnorm(m*n), nr = m)) {
  print(mm)
  # m:  number of rows; n:  number of columns
  lda <- max(1, m)
  ipiv <- min(m, n)
  mmdc = .Fortran("dgetrf", 
                  m = as.integer(m), n = as.integer(n),   # input
                  a = as.double(mm),                      # input/output
                  lda = as.integer(lda),                    # input
                  ipiv = integer(ipiv), info = integer(1))   # output
  return(mmdc)
}
```

> Note the subtle but important difference between `as.integer(4)`,
which is of length 1 and has the value 4, and `integer(4)` which is
an integer vector of 4 zeros.

```{r}
rows = 5
cols = 5
mm = matrix(rnorm(rows*cols), nr = rows)       # input data

dgetrf(rows, cols, mm)
```

## or this other way
```{r}
rows = 4
cols = 4

dgetrf(rows, cols)   ## the matrix is generated at the function args
```

