---
title: "Factorial function in Fortran"
output:
  html_document: default
  html_notebook: default
---

## The Fortran function 
The factorial function resides in the file `facto.f`.
Here is the code.
```
        subroutine facto(n,answer)
c
c simple subroutine to compute factorial
c
        integer n, answer, i
 
        answer = 1
        do 100 i = 2,n
           answer = answer * i
  100   continue 
         
        end
```   

Notice that `facto` is a subroutine not a function, so, we will not need a Fortran wrapper like we did in `fib.f`.

## Compiling the file `facto.f` from the notebook
This could be done from the R prompt directly. In this case, we will link and compile to generate the DLL from the notebook.

Sometimes, the DLL file is locked because it was called by R but not released. If this happens, exit R and try deleting the files again.
```{r results="hold"}
# delete existing files first
file.remove("facto.o")
file.remove("facto.dll")
file.remove("facto.so")

# now, compile
system("R CMD SHLIB facto.f90")

```
The file `facto.dll` should be there now.

## Load the subroutine. Generic OS call
```{r}
# this loads the library in any system, Windows or Linux
dyn.load(paste("facto", .Platform$dynlib.ext, sep = ""))

# or use this
# dyn.load('facto.dll') for Windows only
```

We call this function with:
```{r}
facto <- function(num) {
  out <-  .Fortran("facto", 
                   n = as.integer(num), 
                   answer = as.integer(1))
  return(out$answer)
}

# call the R function wrapper
facto(5)    
```


We can also call the Fortran function multiple times, like this:
```{r}
for(num in 1:10)
  print(.Fortran('facto', 
                 n = as.integer(num), 
                 answer = integer(1))$answer)
```




