---
title: "Fibonacci function in Fortran"
output:
  html_document: default
  html_notebook: default
---

## Calling the Fortran function with the wrapper included
As R can only call Fortran subroutines, sometimes writing a wrap is necessary.
In this case the wrapper with the subroutine is inside the Fortran file.   
This is a sample of the code inside `fib.f90`:
```
    INTEGER FUNCTION fib(n)
      integer n
      integer, parameter :: fib0 = 0, fib1 = 1
      integer back1, back2, i
      select case (n)
        case (:0);      fib = fib0
        case (1);       fib = fib1
        case default
          fib = fib1
          back1 = fib0
          do i = 2, n
            back2 = back1
            back1 = fib
            fib   = back1 + back2
          end do
      end select
    END FUNCTION
    
    SUBROUTINE fib_R_wrapper(n, answer)
      INTEGER n, answer, fib
      EXTERNAL fib
      answer = fib(n)
    END SUBROUTINE
```   

## Compiling the file `fib.f90` from the notebook
This could be done from the R prompt directly. In this case, we will link and compile to generate the DLL from the notebook.

Sometimes, the DLL file is locked because it was called by R but not released. If this happens, exit R and try deleting the files again.
```{r results="hold"}
# delete existing files first
file.remove("fib.o")
file.remove("fib.dll")
file.remove("fib.so")

# now, compile
system("R CMD SHLIB fib.f90")

```

## Load the subroutine. Generic OS call
```{r}
# this loads the library in any system, Windows or Linux
dyn.load(paste("fib", .Platform$dynlib.ext, sep = ""))

# or use this
# dyn.load('fib.dll') for Windows only
```

## Test the function
We call this function with:
```{r}
fib <- function(num) {
  # R function that calls the Fortran function `fib` through
  # a Fortran wrapper.
  out <- .Fortran('fib_R_wrapper',         # `out` is a list and returns the result
                  n=as.integer(num),       # and the parameter that were passed.
                  result = as.integer(1))
  return(out$result)             
}

# call the R function wrapper
fib(10)    
```

## Call the Fortran function with an external Fortran wrapper
In this case we create a Fortran wrapper, separate from the calculating function `fib`. This is the file `fib_R_wrapper.f`:

```
    SUBROUTINE fib_R_wrapper(n, answer)
      INTEGER n, answer, fib
      EXTERNAL fib
      answer = fib(n)
    END SUBROUTINE
```

We use the `.Fortran` function to call the Fortran wrapper.
```{r}
n = 10
.Fortran('fib_R_wrapper', n = as.integer(n), result = integer(1))$result
```

We can also call the Fortran function many times, like this:
```{r}
for(n in 1:10)
  print(.Fortran('fib_R_wrapper', 
                 n = as.integer(n), 
                 result = integer(1))$result)
```

## Unload the Fortran subroutine. Generic OS
```{r}
dyn.unload(paste("fib", .Platform$dynlib.ext, sep = ""))
```



